A comparative analysis of AquaMaps and IUCN species distribution maps

### 2015-12-12 data_explore_2015_data.Rmd
more on manuscript...

Plotting species from q1 (well-matched) (see data_explore_quadmaps.html for plots)
```{r plot_spp_from_q1, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

x <- 20
y <- nrow(spp_list_q1)

spp_names_q1 <- spp_list_q1$sciname[seq(1, y, y/x)]

for (spp in spp_names_q1) { # i = 1
  spp_plot <- plot_rangemap(spp)
  spp_solo <- spp_list %>% filter(sciname == spp)
  text_tag <- sprintf('Overlap: %s%%, area ratio: %s%%', round(spp_solo$sm_perc, 1), round(spp_solo$area_ratio, 1))
  spp_plot <- spp_plot +
    labs(title = sprintf('%s, quadrant 1: %s', spp, text_tag))

  print(spp_plot)

  # ggsave(filename = file.path(dir_fig, sprintf('rangemap_q1_%s.png', str_replace(tolower(spp), ' ', '_'))),
  #        plot = spp_plot, 
  #        width = 8, height = 4, dpi = 200)
}

@@

Plotting species from q2 (distribution-matched)
```{r plot_spp_from_q2, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

x <- 20
y <- nrow(spp_list_q2)

spp_names_q2 <- spp_list_q2$sciname[seq(1, y, y/x)]

for (spp in spp_names_q2) { # i = 1
  spp_plot <- plot_rangemap(spp)
  spp_solo <- spp_list %>% filter(sciname == spp)
  text_tag <- sprintf('Overlap: %s%%, area ratio: %s%%', round(spp_solo$sm_perc, 1), round(spp_solo$area_ratio, 1))
  spp_plot <- spp_plot +
    labs(title = sprintf('%s, quadrant 2: %s', spp, text_tag))
  print(spp_plot)

  # ggsave(filename = file.path(dir_fig, sprintf('rangemap_q2_%s.png', str_replace(tolower(spp), ' ', '_'))),
  #        plot = spp_plot, 
  #        width = 8, height = 4, dpi = 200)
}

@@

Plotting species from q3 (area-matched)
```{r plot_spp_from_q3, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

x <- 20
y <- nrow(spp_list_q3)

spp_names_q3 <- spp_list_q3$sciname[seq(1, y, y/x)]

for (spp in spp_names_q3) { # i = 1
  spp_plot <- plot_rangemap(spp)
  spp_solo <- spp_list %>% filter(sciname == spp)
  text_tag <- sprintf('Overlap: %s%%, area ratio: %s%%', round(spp_solo$sm_perc, 1), round(spp_solo$area_ratio, 1))
  spp_plot <- spp_plot +
    labs(title = sprintf('%s, quadrant 3: %s', spp, text_tag))
  print(spp_plot)

  # ggsave(filename = file.path(dir_fig, sprintf('rangemap_q3_%s.png', str_replace(tolower(spp), ' ', '_'))),
  #        plot = spp_plot, 
  #        width = 8, height = 4, dpi = 200)
}

Plotting species from q4 (poorly matched)
```{r plot_spp_from_q4, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

x <- 20
y <- nrow(spp_list_q4)

spp_names_q4 <- spp_list_q4$sciname[seq(1, y, y/x)]

for (spp in spp_names_q4) { # i = 1
  spp_plot <- plot_rangemap(spp)
  spp_solo <- spp_list %>% filter(sciname == spp)
  text_tag <- sprintf('Overlap: %s%%, area ratio: %s%%', round(spp_solo$sm_perc, 1), round(spp_solo$area_ratio, 1))
  spp_plot <- spp_plot +
    labs(title = sprintf('%s, quadrant 4: %s', spp, text_tag))
  print(spp_plot)

  # ggsave(filename = file.path(dir_fig, sprintf('rangemap_q4_%s.png', str_replace(tolower(spp), ' ', '_'))),
  #        plot = spp_plot, 
  #        width = 8, height = 4, dpi = 200)
}

>>>>

``` {r load data, echo = FALSE, warning = FALSE, message = FALSE}
### use data_explore.Rmd to create this file, if it does not yet exist.

data_file <- 'data/spp_list_w_area_percent_noextinct_2015.csv'
spp_map_pairs <- read.csv(data_file, stringsAsFactors = FALSE) %>%
  filter(!is.na(popn_category)) %>% ### filter out unassessed species
  filter(!is.na(sm_perc))           ### filter out NA maps (usually IUCN maps with no polygons)

### generate other lists

spp_all_file <- 'data/spp_all_2015.csv'
spp_all <- read.csv(spp_all_file, stringsAsFactors = FALSE)

iucn_marine_file <- 'data/spp_iucn_marine.csv'
spp_iucn_marine <- read.csv(iucn_marine_file, stringsAsFactors = FALSE)

iucn_maps_file <- 'data/spp_iucn_maps.csv'
spp_iucn_maps <- read.csv(iucn_maps_file, stringsAsFactors = FALSE) %>%
  filter(spp_group != 'hagfishes') %>%           ### not from IUCN...
  filter(sciname %in% spp_iucn_marine$sciname)   ### ditch freshwater spp

spp_am_file <- 'data/spp_am.csv'
spp_am <- read.csv(spp_am_file, stringsAsFactors = FALSE)

@@

``` {r breakdown by taxa, echo = FALSE}

sciname_lookup <- read.csv('sciname_lookups/spp_iucn_all.csv', stringsAsFactors = FALSE)

iucn_maps_class <- spp_iucn_maps %>% 
  left_join(sciname_lookup %>% 
              select(sciname, class) %>%
              unique(), 
            by = 'sciname')

taxa_iucn_all <- iucn_maps_class[!duplicated(iucn_maps_class$sciname), ] %>%
  filter(!is.na(class)) %>% ### don't count the unknowns
  mutate(class = str_to_title(class),
         n_iucn_tot = n()) %>%        ### length of whole maps list
  group_by(class) %>%
  summarize(n_iucn_class_all  = n(),
            p_iucn_class_all  = round((n_iucn_class_all/first(n_iucn_tot) * 100), 2))

taxa_iucn_solo <- iucn_maps_class[!duplicated(iucn_maps_class$sciname), ] %>%
  filter(!is.na(class)) %>% ### don't count the unknowns
  filter(!sciname %in% spp_am$sciname) %>%  ### filter list to IUCN only
  mutate(class = str_to_title(class),
         n_iucn_solo = n()) %>%       ### length of IUCN only list
  group_by(class) %>%
  summarize(n_iucn_class_solo = n(),
            p_iucn_class_solo = round((n_iucn_class_solo/first(n_iucn_solo) * 100), 2))

taxa_am_all <- spp_am %>%
  filter(!class %in% c('Not assigned', 'null') & !is.na(class)) %>%
  mutate(class = str_to_title(class),
         n_am_tot = n()) %>%        ### length of whole AM maps list
  group_by(class) %>%
  summarize(n_am_class_all  = n(),
            p_am_class_all  = round((n_am_class_all/first(n_am_tot) * 100), 2))

taxa_am_solo <- spp_am %>%
  filter(!class %in% c('Not assigned', 'null') & !is.na(class)) %>%
  filter(!sciname %in% iucn_maps_class$sciname) %>%  ### filter list to AM only
  mutate(class = str_to_title(class),
         n_am_solo = n()) %>%       ### length of AM only list
  group_by(class) %>%
  summarize(n_am_class_solo = n(),
            p_am_class_solo = round((n_am_class_solo/first(n_am_solo) * 100), 2))

taxa_both <- spp_am %>%
  filter(sciname %in% spp_iucn_maps$sciname) %>%
  filter(!class %in% c('Not assigned', 'null')) %>%
  mutate(class = str_to_title(class),
         n_both_tot = n()) %>%
  group_by(class) %>%
  summarize(n_class_both = n(),
            p_class_both = round(n_class_both/first(n_both_tot) * 100, 2))

taxa_df <- taxa_am_all %>%
  full_join(taxa_am_solo,   by = 'class') %>%
  full_join(taxa_iucn_all,  by = 'class') %>%
  full_join(taxa_iucn_solo, by = 'class') %>%
  full_join(taxa_both,      by = 'class') %>%
  filter(!(is.na(n_iucn_class_all) & p_am_class_all < 0.5))

knitr::kable(taxa_df %>%
               select(class, AquaMaps = n_am_class_all, IUCN = n_iucn_class_all, Both = n_class_both))

### bar charts comparing number of species in each class for each data set

taxa_df_long1 <- taxa_df %>%
  select(class, AquaMaps = n_am_class_all, IUCN = n_iucn_class_all, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both)
bar_taxa1 <- ggplot(taxa_df_long1, aes(x = class, y = n_spp)) +
  geom_bar(stat = 'identity') +
  facet_grid(dataset ~ ., scales = 'free_y') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1))
print(bar_taxa1)

taxa_df_long2 <- taxa_df %>%
  select(class, AquaMaps = n_am_class_solo, IUCN = n_iucn_class_solo, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both) %>%
  group_by(class) %>%
  mutate(n_total = sum(n_spp, na.rm = TRUE),
         percent = n_spp/n_total * 100)
bar_taxa2 <- ggplot(taxa_df_long2, aes(x = class, y = percent, fill = dataset)) +
  geom_bar(stat = 'identity') +
  geom_text(aes(label = sprintf('n = %s', n_total), y = .05), hjust = 0, size = 4, angle = 90, hjust = 1) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  labs(x = 'Class', y = 'Percent of mapped species',
       title = 'Distribution of species maps by taxonomic group (class)')
print(bar_taxa2)

@@

``` {r child = 'R/data_explore_pt3.Rmd', eval = FALSE}
# Multiple species maps per quadrant
```

### 2015-12-12 ms_outline.Rmd
more on manuscript...

iucn_maps_file <- 'data/spp_iucn_maps.csv'
spp_iucn_maps <- read.csv(iucn_maps_file, stringsAsFactors = FALSE) %>%
  filter(spp_group != 'hagfishes') %>%           ### not from IUCN...
  filter(sciname %in% spp_iucn_marine$sciname)   ### ditch freshwater spp

@@

```
## Main message

@@

###IUCN

@@

IUCN publishes spatial distribution map sets only for taxonomic groups in which at least 90% of the species within the taxon group have been evaluated.  While this mitigates sampling bias within taxa, it also means that entire taxonomic groups remain unavailable until they have met this threshold of comprehensive assessment.  As of December 2015, IUCN had published species distribution maps for nearly 12,000 marine species across 26 taxonomic groups, though fewer than 4,500 of these have been formally assessed under the IUCN Red List of Threatened Species.

@@

### AquaMaps

@@

### Comparison between AM and IUCN

@@

### Overlap between assessed species

| Map set   | n~total~                | n~Red_List~                                                               |
| :-------: | ----------------------: | ------------------------------------------------------------------------: |
| AquaMaps  | `r nrow(spp_am)`        | `r nrow(spp_am %>% filter(sciname %in% spp_iucn_marine$sciname))`         |
| IUCN      | `r nrow(spp_iucn_maps)` | `r nrow(spp_iucn_maps %>% filter(sciname %in% spp_iucn_marine$sciname))`  |
| Both      | `r nrow(spp_am %>% filter(sciname %in% spp_iucn_maps$sciname))` | `r nrow(spp_map_pairs)`           |

Researchers have two large and useful data sets at their disposal, each with its own utility.  Being able to use both simultaneously makes many more species available, but beware: here there be challenges.

@@

Focus on IUCN red-list evaluated species (incl or not incl DD)? or all species?

@@

Examine the overlapping and non-overlapping species by taxonomic group (order? family?) - are there differences in taxonomic group coverage between the two data sets?

IUCN - translate the taxonomic group into families - is there a way to automate this? http://www.itis.gov/downloads/?
AM - access taxa group information directly from AM species list - order? family?

``` {r breakdown by taxa, echo = FALSE}

sciname_lookup <- read.csv('sciname_lookups/scinames.csv', stringsAsFactors = FALSE)

iucn_maps_class <- spp_iucn_maps %>% 
  ### add sciname list, then jump through hoops to delete dupes
  ### NOTE: this also ditches ones with no sciname match...
  separate(sciname, 'genus', sep = ' ', remove = FALSE, extra = 'drop') %>%
  mutate(genus = tolower(genus)) %>%
  left_join(sciname_lookup, by = 'genus') %>% 
  filter(!(str_detect(tolower(spp_group), 'grass')    & !str_detect(kingdom, 'plant'))) %>%
  filter(!(str_detect(tolower(spp_group), 'mangrove') & !str_detect(kingdom, 'plant'))) %>%
  filter(!(str_detect(tolower(spp_group), 'coral')    & !str_detect(kingdom, 'animal'))) %>%
  filter(!(str_detect(tolower(spp_group), 'reptile')  & !str_detect(kingdom, 'animal'))) %>%
  filter(!(str_detect(tolower(spp_group), 'seabream') & !str_detect(kingdom, 'animal'))) %>%
  filter(!(str_detect(tolower(spp_group), 'wrass')    & !str_detect(kingdom, 'animal'))) %>%
  filter(!(str_detect(tolower(spp_group), 'mamm')     & !str_detect(kingdom, 'animal'))) %>%
  filter(!is.na(phylum)) %>%
  select(-kingdom, -phylum, -family) %>%
  unique()

taxa_iucn_all <- iucn_maps_class[!duplicated(iucn_maps_class$sciname), ] %>%
  filter(!is.na(class)) %>% ### don't count the unknowns
  mutate(class = str_to_title(class),
         n_iucn_tot = n()) %>%        ### length of whole maps list
  group_by(class) %>%
  summarize(n_iucn_class_all  = n(),
            p_iucn_class_all  = round((n_iucn_class_all/first(n_iucn_tot) * 100), 2))

taxa_iucn_solo <- iucn_maps_class[!duplicated(iucn_maps_class$sciname), ] %>%
  filter(!is.na(class)) %>% ### don't count the unknowns
  filter(!sciname %in% spp_am$sciname) %>%  ### filter list to IUCN only
  mutate(class = str_to_title(class),
         n_iucn_solo = n()) %>%       ### length of IUCN only list
  group_by(class) %>%
  summarize(n_iucn_class_solo = n(),
            p_iucn_class_solo = round((n_iucn_class_solo/first(n_iucn_solo) * 100), 2))

taxa_am_all <- spp_am %>%
  filter(!class %in% c('Not assigned', 'null') & !is.na(class)) %>%
  mutate(class = str_to_title(class),
         n_am_tot = n()) %>%        ### length of whole AM maps list
  group_by(class) %>%
  summarize(n_am_class_all  = n(),
            p_am_class_all  = round((n_am_class_all/first(n_am_tot) * 100), 2))

taxa_am_solo <- spp_am %>%
  filter(!class %in% c('Not assigned', 'null') & !is.na(class)) %>%
  filter(!sciname %in% iucn_maps_class$sciname) %>%  ### filter list to AM only
  mutate(class = str_to_title(class),
         n_am_solo = n()) %>%       ### length of AM only list
  group_by(class) %>%
  summarize(n_am_class_solo = n(),
            p_am_class_solo = round((n_am_class_solo/first(n_am_solo) * 100), 2))

taxa_both <- spp_am %>%
  filter(sciname %in% spp_iucn_maps$sciname) %>%
  filter(!class %in% c('Not assigned', 'null')) %>%
  mutate(class = str_to_title(class),
         n_both_tot = n()) %>%
  group_by(class) %>%
  summarize(n_class_both = n(),
            p_class_both = round(n_class_both/first(n_both_tot) * 100, 2))

taxa_df <- taxa_am_all %>%
  full_join(taxa_am_solo,   by = 'class') %>%
  full_join(taxa_iucn_all,  by = 'class') %>%
  full_join(taxa_iucn_solo, by = 'class') %>%
  full_join(taxa_both,      by = 'class') %>%
  filter(!(is.na(n_iucn_class_all) & p_am_class_all < 0.5))

knitr::kable(taxa_df %>%
               select(class, am_perc = p_am_class_all, iucn_perc = p_iucn_class_all, both_perc = p_class_both))
```

@@

Using predifined species groups by the IUCN, it is clear to see that tropical species - specifically wrasses, corals, blennies and groupers, make up the majority of species found in both. This is driven more by the IUCN dataset since the number of species with spatial data is directly attributed to the species that have assessments. ___This could be shown using the n = XXX on the species group quadrant breakdown chart too.___

_Total species occurrences for each data set: all AM, all IUCN, and overlap (one? the other? occurrences of both?)_

@@

@@

* For IUCN data, we defined "presence" within a cell as any non-zero overlap with a range map polygon.
* For AquaMaps data, we defined presence as any cell with non-zero probability ___(or RES?)___.  
* In both cases, these criteria for presence will maximize the number of cells included.  Using these criteria, we found that for 
`r round(nrow(spp_map_pairs %>% filter(sm_range == 'AM')) / nrow(spp_map_pairs) * 100, 1)`%
of species, the IUCN distribution maps indicated a larger species range than the range indicated by AquaMaps.

@@

``` {r calc medians and quadrant counts, echo = FALSE}
dist_median <- round(median(spp_map_pairs$sm_perc,    na.rm = TRUE), 1)
area_median <- round(median(spp_map_pairs$area_ratio, na.rm = TRUE), 1)
q1_count <- sum(spp_map_pairs$sm_perc > dist_median  & spp_map_pairs$area_ratio >  area_median)
q2_count <- sum(spp_map_pairs$sm_perc <= dist_median & spp_map_pairs$area_ratio >  area_median)
q3_count <- sum(spp_map_pairs$sm_perc > dist_median  & spp_map_pairs$area_ratio <= area_median)
q4_count <- sum(spp_map_pairs$sm_perc <= dist_median & spp_map_pairs$area_ratio <= area_median)
```

| Description              |   Distribution match |      Area match      |
| :----------------------: | :------------------: | :------------------: |
| q1: Well-aligned         |  > `r dist_median` % |  > `r area_median` % |
| q2: Distribution-aligned |  < `r dist_median` % |  > `r area_median` % |
| q3: Area-aligned         |  > `r dist_median` % |  < `r area_median` % |
| q4: Poorly-aligned       |  < `r dist_median` % |  < `r area_median` % |

>>>>

iucn_maps_file <- 'data/spp_iucn_maps.csv'
spp_iucn_maps <- read.csv(iucn_maps_file, stringsAsFactors = FALSE) %>%
  rename(sciname = binomial) %>%
  filter(spp_group != 'hagfishes') %>%           ### not from IUCN...
  filter(sciname %in% spp_iucn_marine$sciname)   ### ditch freshwater spp

@@

```

## Main message

@@

## IUCN species distribution maps

@@

IUCN publishes spatial distribution map sets only for taxonomic groups whjich have been "comprehensively assessed," i.e. in which at least 90% of the species within the taxon group have been evaluated.  While this mitigates sampling bias within taxa, it also means that entire taxonomic groups remain unavailable until they have met this threshold of comprehensive assessment.  As of December 2015, IUCN had published species distribution maps for nearly 12,000 marine species across 26 taxonomic groups, though fewer than 4,500 of these have been formally assessed under the IUCN Red List of Threatened Species.

@@

## AquaMaps standardized range maps of marine species

@@

## Comparison between AM and IUCN

@@

@@

## Overlap between assessed species

Due to differences in both methodology and intent, there are significant differences in the coverage afforded by the species distribution maps published by AquaMaps and those published by the IUCN.  This is immediately clear from an examination of the number of species maps offered by each source, and the small proportion of species with maps from both sources.

| Map set   | n~total~                | n~Red_List~                                                       |
| :-------: | ----------------------: | :---------------------------------------------------------------: | 
| AquaMaps  | `r nrow(spp_am)`        | `r nrow(spp_am %>% filter(sciname %in% spp_iucn_marine$sciname))` | 
| IUCN      | `r nrow(spp_iucn_maps)` | all                                                               | 
| Both      | `r nrow(spp_am %>% filter(sciname %in% spp_iucn_maps$sciname))` | all                       |

___Should we break it down by "reviewed" Aquamaps as well?___

The AquaMaps species list includes all species available as of December 2015.  The IUCN species list includes all species maps available directly from the IUCN Red List website (http://www.iucnredlist.org/technical-documents/spatial-data) as of December 2015, filtered to include only species with habitat listed as "marine."  For the purposes of this analysis, we did not include IUCN range maps for bird species, as those range maps are hosted separately by BirdLife International.

@@

![](figures/bar_by_taxa_all.png)

![](figures/bar_by_taxa_assessed.png)

As IUCN only releases map sets when a taxonomic group (on the scale of order or family) has been comprehensively assessed, spatial data for many taxonomic classes are not available, and within a class, the assessed sub-groups may not represent the entire class.  For example, as of this writing, IUCN has released no spatial data for class Elasmobranchii (cartilaginous fishes including sharks and rays); and while IUCN offers a large number of maps within class Actinopterygii (ray-finned bony fishes), the available maps include only a few primarily tropical taxonomic sub-groups, such as wrasses, damselfish, butterflyfish, tunas, and billfishes.  However, IUCN's criterion of comprehensive assessment greatly reduces the risk of sample bias within the bounds of the assessed taxonomic groups.

The release of AquaMaps distribution maps is not limited in the same way, and maps are available across a much larger range of taxonomic classes; however, there is no guarantee that the list of species included within each class is a representative cross section of the entire class.  ___How do they choose their species to assess?___

@@

Using predefined species groups by the IUCN, it is clear to see that tropical species - specifically wrasses, corals, blennies and groupers, make up the majority of species found in both. This is driven more by the IUCN dataset since the number of species with spatial data is directly attributed to the species that have assessments. ___This could be shown using the n = XXX on the species group quadrant breakdown chart too.___

___Total species occurrences for each data set: all AM, all IUCN, and overlap (one? the other? occurrences of both?)___

@@

Each data set offers spatial distribution information for large numbers of species.  However, the data sets vary in terms of taxonomic coverage and regional coverage.  For spatial assessments of biodiversity, the choice of one data set over the other is likely to create significantly different results.  For studies confined to a narrow range of taxa or to a narrow spatial scale, one data set may offer an advantage over the other in the number of species maps available.  For global scale biodiversity studies, however, the selection of one data set over the other will entail tradeoffs in spatial coverage, taxonomic breadth, and taxonomic depth. ___find a reference that describes what makes a "good" data set for global biodiversity, e.g. OHI___

@@

* For IUCN data, we defined "presence" within a cell as any non-zero overlap with a range map polygon.
* For AquaMaps data, we defined presence as any cell with non-zero probability ___(or RES?)___.  
* In both cases, these criteria for presence will maximize the number of cells included.

Using these criteria, we found that for `r round(nrow(spp_map_pairs %>% filter(sm_range == 'AM')) / nrow(spp_map_pairs) * 100, 1)`% of species, the IUCN distribution maps indicated a larger species range than the range indicated by AquaMaps.

@@

``` {r calc medians and quadrant counts, echo = FALSE}
dist_median <- round(median(spp_map_pairs$sm_perc,    na.rm = TRUE), 1)
area_median <- round(median(spp_map_pairs$area_ratio, na.rm = TRUE), 1)
q1_count <- sum(spp_map_pairs$sm_perc >= dist_median & spp_map_pairs$area_ratio >= area_median)
q2_count <- sum(spp_map_pairs$sm_perc <  dist_median & spp_map_pairs$area_ratio >= area_median)
q3_count <- sum(spp_map_pairs$sm_perc >= dist_median & spp_map_pairs$area_ratio <  area_median)
q4_count <- sum(spp_map_pairs$sm_perc <  dist_median & spp_map_pairs$area_ratio <  area_median)
```

| Description              |  Distribution match  |     Area match       |
| :----------------------: | :------------------: | :------------------: |
| q1: Well-aligned         | >= `r dist_median` % | >= `r area_median` % |
| q2: Distribution-aligned |  < `r dist_median` % | >= `r area_median` % |
| q3: Area-aligned         | >= `r dist_median` % |  < `r area_median` % |
| q4: Poorly-aligned       |  < `r dist_median` % |  < `r area_median` % |
